---
title: "Differential expression analysis"
output: html_document
editor_options: 
  chunk_output_type: console
---

## Setup

Load packages.

```{r packages}
library(limma)
library(edgeR)
```

Import counts.

```{r import}
rawData <- read.table("data/counts.txt",
                      header = TRUE,
                      stringsAsFactors = FALSE)
dim(rawData)
genes <- subset(rawData, select = Geneid:Length)
counts <- rawData[, 7:18]
rownames(counts) <- genes$Geneid
colnames(counts) <- gsub("\\.*bam\\.*", "", colnames(counts))
group <- c(rep("mutant", 6), rep("wildtype", 6))
group <- factor(group, levels = c("wildtype", "mutant"))
x <- DGEList(counts = counts,
             group = group,
             genes = genes)
class(x)
```

## Process features

<!-- Ex 1: log2CPM -->

Calculate log2 counts per million (log2cpm).

```{r cpm}
log2cpm <- cpm(x, log = TRUE)
```

Plot density of expression values for each sample.

```{r plotDensities}
plotDensities(log2cpm, group = group, main = "Raw")
```

<!-- Ex 2: Filter -->

Only keep features which have at least 10 counts in at least 4 wildtype or 4
mutant samples.

```{r keep}
keep <- filterByExpr(x, group = group)
sum(keep)
x <- x[keep, ]
```

Re-calculate log2cpm and re-plot densities.

```{r plotDensitiesFiltered}
log2cpm <- cpm(x, log = TRUE)
plotDensities(log2cpm, group = group, main = "Filtered")
```

<!-- Ex 3: Normalize -->

Normalize the samples, re-calculate log2cpm, and re-plot densities.

```{r normalize}
x <- calcNormFactors(x)
log2cpm <- cpm(x, log = TRUE)
plotDensities(log2cpm, group = group, main = "Normalized")
```

## Process samples

<!-- Ex 1: SNF2 -->

Confirm that the mutant samples are null for SNF2.

```{r snf2}
barplot(log2cpm["YOR290C", ], main = "SNF2")
```

<!-- Ex 2: PCA -->

Perform PCA.

```{r pca}
plotMDS(log2cpm, gene.selection = "common")
```

Remove outlier sample.

```{r}
x <- x[, colnames(x) != "mutant.06"]
dim(x)
```

<!-- Ex 3: PCA 2 -->

Re-calculate log2cpm and re-perform PCA.

```{r pca2}
log2cpm <- cpm(x, log = TRUE)
plotMDS(log2cpm, gene.selection = "common")
```

## Model

$$ Y = \beta_{0} + \beta_{mutant} + \epsilon $$

```{r design}
design <- model.matrix(~x$samples$group)
```

```{r voom}
v <- voom(x, design, plot = TRUE)
```

```{r fit}
fit <- lmFit(v, design)
fit <- eBayes(fit)
```

## Explore results

<!-- Ex 1: topTable -->

Count number of differentially expressed features.

```{r number-of-de-genes}
summary(decideTests(fit))
```

View top 10 differentially expressed genes.

```{r}
topTable(fit, coef = 2)
```

Create a barplot of the top DE feature:

```{r}
barplot(log2cpm["YML123C", ], las = 2, cex.names = 0.75)
```

<!-- Ex 2: assessment -->

Visualize p-value distribution.

```{r p-values}
hist(fit$p.value[, 2], main = "p-value distribution")
```

Visualize residual variation versus magnitude of expression.

```{r residuals}
plotSA(fit)
```

<!-- Ex 3: volcano -->

Create a volcano plot.

```{r volcano}
volcanoplot(fit, coef = 2, highlight = 5, names = fit$genes$Geneid)
```
